National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs

### 2021-02-12 Manuscript_Bayesian.Rmd
fixed bib

Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life Force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

Beliefs & \_Skeptic\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\

@@

### Belief in a Spirit/Life Force and/or a God  
We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life Force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer.

## Statistical Analyses
Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (*N* = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years. This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years. We estimated all statitiscal models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors. We include the full reports in the Supplement. The REML estimates and the Bayesian estimates converged to within three decimal places.  We include model equations and priors at the end of this article. Here we report the results of the Bayesian estimator.

@@

This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g. retirement).  Results were insensitive to the choice set. Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details). In the fully saturated model, it appeared that those skeptical of a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years. Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking. We discuss this point in more detail in the supplement, and mention it here to caution researchers from "everything-but-the-kitchen-sink" approaches to regression modelling.

@@

  - The intercept, representing the expected personal well-being among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI [ 7.111,  7.167], pd = 100.00%).  
  - The effect of years is positive is reliable (posterior median =  7.138, HDI [ 7.111,  7.167], pd  100.00%).  We infer that personal well-being increased in the baseline skeptic population between 2010 and 2018.
  - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median =  -0.059, HDI [-0.081, -0.036], pd 100.00%). We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
  - The effect of beliefs in *God And Spirit*  is not  negative (posterior median = -0.060, HDI [-0.086, -0.034], pd = 100.00%) . We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
  - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.001 HDI [-0.036,  0.035], pd = 51.10%). We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.
  - The effect of Age in Decades (C) is   0.072 | [ 0.060,  0.083] | 100.00% |
  - The effect of Maori Ethnicity is  -0.380 | [-0.420, -0.341] | 100.00% | 
  - The effect of Pacific Ethnicity is  -0.384 | [-0.464, -0.312] | 100.00% |
  - The effect of Asian Ethnicity is -0.093 | [-0.165, -0.023] |  98.28% |  
  - The effect of Male gender is  -0.060 | [-0.094, -0.026] |  99.90% |

@@

The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.011, HDI [ 0.003,  0.018], pd = 99.02%). We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.007, 0.021], pd = 99.98%). We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
  - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.004, HDI [-0.009,  0.015], pd =  68.73%).  We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than skeptics grew.

@@

Within this model: 
  - The intercept, representing the expected life satisfaction among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI [ 5.267,  5.306], pd = 100.00%).
  - The effect of years is positive is reliable (posterior median =  0.008, HDI [ 0.003, 0.011], pd = 99.75%).  We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
  -The effect of beliefs in *Spirit Excludes God*  is not reliable (posterior median =   0.007, HDI [-0.010,  0.023], pd = 75.33%). We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
  - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.034, HDI [ 0.015, 0.053], pd = 99.83%). We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
  - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.057, HDI [ 0.030, 0.082], pd = 99.95%). We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with skeptics.
  - The effect of Age in Decades (C) is reliably positive (posterior median = 0.079, hDI [ 0.071,  0.087] pd = 100.00%).
  - The effect of Maori Ethnicity is reliably negative (posterior median = -0.109 | [-0.137, -0.082], pd = 100.00%).
  - The effect of Pacific Ethnicity is reliably negative (posterior median = -0.153 | [-0.210, -0.103], pd = 100.00%).
  - The effect of Asian Ethnicity is not reliable (posterior median = -0.047, [-0.098,  0.004], pd = 93.08%).
  - The effect of Male gender is reliabily negative (posterior median = -0.161, [-0.184, -0.138], pd = 100.00%)

@@

```{r  fig.cap = "Life satisfaction prediction graph", echo = FALSE,out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted.png")
```
  - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.011, HDI [ 0.006, 0.016], pd = 99.95%). We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
  - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.009, HDI [ 0.004, 0.014], pd = 99.72%). We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.003, HDI [-0.012, 0.006], pd = 72.28%). We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to skeptics.

@@

1. At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.93 | [6.88, 6.98]) and lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 | [6.87, 6.95]) in comparison both to the skeptical population who neither believed in a God nor Spirit/ Life Force ($\hat{y}$ = 7.04 | [6.99, 7.10]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 7.03 | [6.95, 7.10]). (See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =   5.20 | [5.16, 5.23]). Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God  ($\hat{y}$ =   5.23 | [5.21, 5.26]) in comparison both with the skeptical population who neither believed in a God nor a Life Force ($\hat{y}$ = 5.25 | [5.21, 5.29]). Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.32 | [5.26, 5.38]) (see Figure 3)

2. However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =   7.19$ [7.15, 7.23], $\Delta \hat{y} = 0.26$). Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown to ($\hat{y} = 7.20$ [7.17, 7.24], $\Delta \hat{y} = 0.29$). By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} =  7.21$ [7.17, 7.25], $\Delta \hat{y} = 0.17$). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  7.22$ [7.16, 7.28] $\Delta \hat{y} = 0.19$). In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life Force beliefs was about 30% greater than it was among those who lacked such Spirit/Life Force (See Figure 2).  Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.  In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.37$ [5.34, 5.40], $\Delta \hat{y} = 0.17$). Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to ($\hat{y} = 5.39$  [5.36, 5.41], $\Delta \hat{y} = 0.19$).By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =   5.32$ [5.29, 5.35], $\Delta \hat{y} = 0.07$). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.36$ [5.31, 5.41], $\Delta \hat{y} = 0.04$). In practical terms, the rate of increase in life satisfaction among those with Spirit/Life Force beliefs over three times greater than among those that lacked such beliefs (See Figure 3).

@@

Two fundamental question arise from the patterns we observe in this study. The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics? During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [@greaves2015regional]. Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [@sibley2012faith]. Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [@norenzayan2006belief]. Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster. On the other hand, such exogenous influences might have adversely affected those with Spirit/Life Force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean. Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations.

@@

# Appendix 
## Model Equations
### Priors for the Personal Wellbeing model

Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes are as follows

$$
\begin{aligned}
\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
\begin{bmatrix}
\alpha_j \\
\beta_j
\end{bmatrix}
\sim 
\text{MVNormal}
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}
,\boldsymbol{S} \nonumber
\end{pmatrix}\\
\boldsymbol{S} = 
\begin{pmatrix}
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix}\\
\sigma_{\alpha_j} \sim 
\text{HalfCauchy}(0,2.5) \nonumber\\
\sigma_{\beta_j} \sim 
\text{HalfCauchy}(0,2.5) \nonumber\\
\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
\end{aligned}
$$

### Priors for the Life Satisfaction model

Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:

$$
\begin{aligned}
\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
\begin{bmatrix}
\alpha_j \\
\beta_j
\end{bmatrix}
\sim 
\text{MVNormal}
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}
,\boldsymbol{S} \nonumber
\end{pmatrix}\\
\boldsymbol{S} = 
\begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix}\\
\sigma_{\alpha_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\sigma_{\beta_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
\end{aligned}
$$
\newpage

### Deterministic model

The deterministic portion of both the Personal Wellbeing and the Life Satisfaction models is as follows: 
$$
\begin{aligned}
& \quad  \boldsymbol{b} = \\
& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\  + \\
& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
&\quad \beta_{4}(\operatorname{Age in Decades (C)})\  +\\ 
& \quad  \beta_{5}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
&\quad \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
& \quad  \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
& \quad \beta_{8}(\operatorname{Male} )+ \\ 
& \quad \beta_{9}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
& \quad \beta_{10}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{11}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
\end{aligned}
$$
\newpage

>>>>

Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and dis0beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual and religious beliefs and dis-beliefs are as follows: (1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: Those who believe in a both God and a Spirit or Life Force. Clearly such an approach to categorising religion and spirituality cannot simultaneously address many interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of religious and spiritual beliefs and dis-beliefs are straightforwardly disentangled and measured within people, and their longitudinal associations with reliable well-being indicators are systematically quantified Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

Beliefs & \_Disbeliefs\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\

@@

### Belief in a Spirit/Life Force and/or a God  
We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as 1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: those who believe in a both God and a Spirit or Life Force.

## Statistical Analyses
Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (*N* = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years. This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years. We estimated all statitiscal models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors. We include the full reports in the Supplement. The REML estimates and the Bayesian estimates converged to within three decimal places.  We include model equations and priors at the end of this article. Here we report the results of the Bayesian estimator.

@@

This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g. retirement).  Results were insensitive to the choice set. Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details). In the fully saturated model, it appeared that those disbelieving in a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years. Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking. We discuss this point in more detail in the supplement, and mention it here to caution researchers from "everything-but-the-kitchen-sink" approaches to regression modelling.

@@

  - The intercept, representing the expected personal well-being among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI [ 7.111,  7.167], pd = 100.00%).  
  - The effect of years is positive is reliable (posterior median =  7.138, HDI [ 7.111,  7.167], pd  100.00%).  We infer that personal well-being increased in the baseline disbelieving population between 2010 and 2018.
  - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median =  -0.059, HDI [-0.081, -0.036], pd 100.00%). We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to disbelievers.
  - The effect of beliefs in *God And Spirit*  is not  negative (posterior median = -0.060, HDI [-0.086, -0.034], pd = 100.00%) . We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to disbelievers.
  - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.001 HDI [-0.036,  0.035], pd = 51.10%). We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to the disbelieving population:
  - The effect of Age in Decades (C) is   0.072 | [ 0.060,  0.083] | 100.00% |
  - The effect of Maori Ethnicity is  -0.380 | [-0.420, -0.341] | 100.00% | 
  - The effect of Pacific Ethnicity is  -0.384 | [-0.464, -0.312] | 100.00% |
  - The effect of Asian Ethnicity is -0.093 | [-0.165, -0.023] |  98.28% |  
  - The effect of Male gender is  -0.060 | [-0.094, -0.026] |  99.90% |

@@

The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.011, HDI [ 0.003,  0.018], pd = 99.02%). We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.007, 0.021], pd = 99.98%). We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
  - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.004, HDI [-0.009,  0.015], pd =  68.73%).  We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than disbelievers grew.

@@

Within this model: 
  - The intercept, representing the expected life satisfaction among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI [ 5.267,  5.306], pd = 100.00%).
  - The effect of years is positive is reliable (posterior median =  0.008, HDI [ 0.003, 0.011], pd = 99.75%).  We infer that life satisfaction increased in the disbelieving population between 2010 and 2018.
  -The effect of beliefs in *Spirit Excludes God*  is not reliable (posterior median =   0.007, HDI [-0.010,  0.023], pd = 75.33%). We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with disbelievers.
  - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.034, HDI [ 0.015, 0.053], pd = 99.83%). We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with disbelievers.
  - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.057, HDI [ 0.030, 0.082], pd = 99.95%). We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with disbelievers.
  - The effect of Age in Decades (C) is reliably positive (posterior median = 0.079, hDI [ 0.071,  0.087] pd = 100.00%).
  - The effect of Maori Ethnicity is reliably negative (posterior median = -0.109 | [-0.137, -0.082], pd = 100.00%).
  - The effect of Pacific Ethnicity is reliably negative (posterior median = -0.153 | [-0.210, -0.103], pd = 100.00%).
  - The effect of Asian Ethnicity is not reliable (posterior median = -0.047, [-0.098,  0.004], pd = 93.08%).
  - The effect of Male gender is reliabily negative (posterior median = -0.161, [-0.184, -0.138], pd = 100.00%)

@@

```{r  fig.cap = "Life satisfaction prediction graph", echo = FALSE,out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted.png")
```
  - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.011, HDI [ 0.006, 0.016], pd = 99.95%). We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
  - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.009, HDI [ 0.004, 0.014], pd = 99.72%). We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.003, HDI [-0.012, 0.006], pd = 72.28%). We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to disbelievers.

@@

1. At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.93 | [6.88, 6.98]) and lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 | [6.87, 6.95]) in comparison both to the population who neither believed in a God nor Spirit/ Life Force ($\hat{y}$ = 7.04 | [6.99, 7.10]) and to the population that believed in a God but was disbelieving in a Spirit or Life Force ($\hat{y}$ 7.03 | [6.95, 7.10]). (See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =   5.20 | [5.16, 5.23]). Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God  ($\hat{y}$ =   5.23 | [5.21, 5.26]) in comparison both with the disbelieving population who neither believed in a God nor a Life Force ($\hat{y}$ = 5.25 | [5.21, 5.29]). Life satisfaction was highest among the population that believed in a God but was disbelieving in a Spirit or Life Force ($\hat{y}$ 5.32 | [5.26, 5.38]) (see Figure 3)

2. However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =   7.19$ [7.15, 7.23], $\Delta \hat{y} = 0.26$). Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown to ($\hat{y} = 7.20$ [7.17, 7.24], $\Delta \hat{y} = 0.29$). By contrast, personal well-being among the disbelieving population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} =  7.21$ [7.17, 7.25], $\Delta \hat{y} = 0.17$). By the same token, personal well-being among those who believe in a God but are disbelieving a Spirit/Life Force grew less steeply $\hat{y} =  7.22$ [7.16, 7.28] $\Delta \hat{y} = 0.19$). In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life Force beliefs was about 30% greater than it was among those who lacked such Spirit/Life Force (See Figure 2).  Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.  In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.37$ [5.34, 5.40], $\Delta \hat{y} = 0.17$). Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to ($\hat{y} = 5.39$  [5.36, 5.41], $\Delta \hat{y} = 0.19$).By contrast, life satisfaction among the population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =   5.32$ [5.29, 5.35], $\Delta \hat{y} = 0.07$). Relatedly, life satisfaction in the population that believe in God but are disbelieving in a Spirit/Life Force grew less steeply $\hat{y} =  5.36$ [5.31, 5.41], $\Delta \hat{y} = 0.04$). In practical terms, the rate of increase in life satisfaction among those with Spirit/Life Force beliefs over three times greater than among those that lacked such beliefs (See Figure 3).

@@

Two fundamental question arise from the patterns we observe in this study. The first question is why those who believed in Spirit or Life Force  were expected to be initially lower relative the populations that did not believe in a Spirit or Life Force? During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [@greaves2015regional]. Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [@sibley2012faith]. Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [@norenzayan2006belief]. Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster. On the other hand, such exogenous influences might have adversely affected those with Spirit/Life Force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean. Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations.

@@
